---
title: 'Regression Models Course Project: Analysis of Mileage and Transmission Type'
author: "juradom"
date: "April 17, 2015"
output: pdf_document
---
####Executive Summary
The goal of this analysis is to answer the following two questions leveraging the mtcars dataset within R:

* Is an automatic or manual transmission better for MPG
* Quantify the MPG difference between automatic and manual transmissions"

####Exploratory Analysis
```{r echo=FALSE}
library(datasets)
data(mtcars)

##factorizing variables
mtcars$am <-factor(mtcars$am,
                   labels=c("automatic","manual"))
mtcars$vs <-factor(mtcars$vs,
                   labels=c("V","S"))

mpg.manual <- mtcars$mpg[mtcars$am=="manual"]
mpg.auto <- mtcars$mpg[mtcars$am=="automatic"]
## determine mean
mean.manual <- mean(mpg.manual)
sd.manual <- sd(mpg.manual)

mean.auto <- mean(mpg.auto)
sd.auto <- sd(mpg.auto)

## determine variance
var.manual <- var(mtcars$mpg[mtcars$am=="manual"])
var.auto <- var(mtcars$mpg[mtcars$am=="automatic"])

## test for normality
shapiro.manual <- shapiro.test(mpg.manual)
shapiro.auto <- shapiro.test(mpg.auto)

```
As described in the R documentation, the mtcars dataset is derived from a 1974 Motor Trend magazine. 
 
* **mtcars Structure**:
    - This data contains 32 observation of 11 variables.
    - 19 of the observations are automatic transmission; 13 are manual  
    - _Appendix A_ contains other structural info about the mtcars dataset.
* Research note: Around the time of the Motor Trend study, a 1975 EPA document was published indicating that weight and engine displacement are the most important factors in fuel economy (http://nepis.epa.gov/Exe/ZyPURL.cgi?Dockey=9100S2LD.txt) 
* **Mean Values**: The average miles per gallon of the manual sample as: **`r round(mean.manual, digits=2)`** and the automatic transmission mean as: **`r round(mean.auto, digits=2)`**.
* **Variances**: The variances for manual and automatic transmission are: **`r round(var.manual, digits=2)`** and **`r round(var.auto,digits=2) `**, respectively.  I suspect the variance issue is due to the small sample size so I will assume they are the same.
* **Normality**: _Appendix B_ shows two density plots that appear to be normal but to be sure I performed a Shapiro test for normality.  The p-value for the manual plot is: **`r round(shapiro.manual$p.value,digits=2) `**.  The p-value for the automatic plot is: **`r round(shapiro.auto$p.value,digits=2) `**.  Since null hypothesis in the Shapiro tests **against** the assumption of normalty, we cannot reject the null hypothesis because both p-values are > .05.  
 
####Regression Analysis

#####**_Means Testing_**  
```{r}
#am.wilcox.test <- suppressWarnings(wilcox.test(mtcars$mpg~mtcars$am))
am.t.test <- t.test(mtcars$mpg~mtcars$am)
```
The p-value for the t-test is: **`r round(am.t.test$p.value, digits=4)`**, therefore, we can reject the null hypothesis and determine that the two population means are different.  

#####**_Interpreting Coefficients And Finding The Best Fitting Model_**  
_Appendix C_ summarizes the base model **lm(mpg~.)**.  This model can explain 81% of the variance in mpg.  None of the p-values are less than 0.05 we can't reject Ho which means none of the variables make a significant contribution.  

Alternatively we can still find the best fitting model by performing a Stepwise Algorithm.  _Appendix D_ shows the output and reveals that with this model ~83% of the variance in mpg can be explained using this model.  The p-values are all < 0.05 which means that wt, qsec, and am are all significant variables.

```{r results='hide'}
fit <- lm(mpg~., data=mtcars)
model.step <- step(fit)
best.step.model <- summary(model.step)$call
```
We conclude that the best formula is:
```{r echo=FALSE}
best.step.model
```
#####**_Finding Interactions_**
Below is a test to see if any interactions are significant.
```{r}
am.wt <- summary(lm(formula = mpg ~ wt + qsec + am + am:wt, data = mtcars))$coef[5,4]
am.qsec <- summary(lm(formula = mpg ~ wt + qsec + am + am:qsec, data = mtcars))$coef[5,4]
qsec.wt <- summary(lm(formula = mpg ~ wt + qsec + am + qsec:wt, data = mtcars))$coef[5,4]
best.fit.model <- summary(lm(formula = mpg ~ wt + qsec + am + am:wt, data = mtcars))$call
```
Based on the different combinations, it appears that the am:wt interaction with a p-value of: `r round(am.wt, digits=4)`. So our best fitted model is: **summary(lm(formula = mpg ~ wt + qsec + am + am:wt, data = mtcars))$call**

#####**_ANOVA_**
```{r}
anova(fit,best.fit.model)
```

####Conclusions
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

\pagebreak

####Appendix A: Structural Analysis of mtcars

```{r}

library(ggplot2)

str(mtcars)
base.plot <- qplot(am, mpg, data=mtcars) 
mpg.am.coef <- coef(lm(mpg~am, data=mtcars))
base.plot + geom_abline(intercept=mpg.am.coef[1], slope=mpg.am.coef[2])

```    

####Appendix B: Testing Normality  

```{r echo=FALSE}
par(mfrow = c(1,2))
##manual histogram
n.man <- NROW(mtcars$am[mtcars$am=="manual"])
hist(mtcars$mpg[mtcars$am=="manual"],
     breaks=n.man,
     xlim = c(10,40),
     xlab="MPG",
     prob=TRUE,
     main="Manual Trans Density")
lines(density(mtcars$mpg[mtcars$am=="manual"]))  

#auto histogram
n.auto <- NROW(mtcars$am[mtcars$am=="automatic"])
hist(mtcars$mpg[mtcars$am=="automatic"],
     breaks=n.auto,
     xlim = c(5,40),
     xlab="MPG",
     prob=TRUE,
     main="Automatic Trans Density")
lines(density(mtcars$mpg[mtcars$am=="automatic"]))

 
```

####Appendix C: Summary Of The Base Model
```{r}
summary(fit)
```
####Appendix D: Summary Of The Best Model
```{r}
summary(model.step)
```
